Evaluating infectious disease forecasts with allocation scoring rules

### 2023-11-03 alloscore_manuscript/alloscore-application.Rnw
some cleanup of intro paragraph 2

Here we pursue one possible tether between multivariate forecasts and policy: the allocation of limited medical resources in response to COVID-19 hospitalizations in various regions so as to minimize expected unmet need. Given probabilistic forecasts of hospitalizations in each region, we formulate an allocation algorithm following techniques developed in operations research. We then score forecasts according to how much unmet need their associated allocations would have allowed. We illustrate this scheme with quantile forecasts of COVID-19 hospitalizations in the US at the state level that are recorded in the COVID-19 Forecast Hub, with the goal of determining the allocation of a hypothetical limited resource across the states. The forecast skill ranking given by this allocation scoring rule can vary substantially from the ranking given by the weighted interval score that is often used to score quantile forecasts, especially during surges in hospitalizations such as in late 2021 as the Omicron wave began. We see this as strong evidence that the allocation scoring rule detects forecast value that is missed by traditional accuracy measures and that the general strategy of designing scoring rules that are directly linked to policy performance is a promising research direction for epidemic forecast evaluation.

@@

Infectious disease forecasting models have emerged as important tools in public health. Predictions of disease dynamics have been used to inform decision making about a wide variety of measures to reduce disease spread and/or mitigate the severity of disease outcomes. For example, estimates of expected onset of flu season have been used to aid national vaccination strategies \citep{igboh2023timing}, and forecasts of Ebola dynamics have been used to allocate surveillance resources \citep{meltzer2014estimating, rainisch2015regional}. \cite{bertsimas2021predictionsCOVID} developed tools to inform decision making from infectious disease forecasts, which have been used to inform allocation of limited medical supplies such as ventilators, ICU capacity planning, and vaccine distribution strategy. Models developed by \cite{fox_real-time_2022} have been used to inform resource and care site planning, as well as community guidelines for masking, traveling, dining and shopping (University of Texas, 2022) \nocite{utnews2022}. In April of 2022, the Centers for Disease Control and Prevention (CDC) announced the launch of the Center for Forecasting and Outbreak Analytics (CFA) to translate disease forecasts into decision making (CDC, 2022)\nocite{cdc2022cfa}, indicating that this has been identified as an important direction at the highest levels of government public health response. The value of infectious disease forecasts has typically been measured by how closely they predict disease outcomes such as cases, hospitalizations or deaths using, for example, root mean square error (RMSE) \citep{papastefanopoulos2020covid} or weighted interval score (WIS) \citep{bracher2021evaluating}, however, recently authors have been calling for evaluating forecasts through their impact on policy \citep{marshall2023predictions}.

In decision making settings where it is possible to quantify the utility or loss associated with a particular action, standard tools of decision theory provide a procedure for developing forecast scoring rules that measure the value of forecasts through the quality of the decisions that they lead to. We give an overview of these procedures in Section \ref{sec:methods.decisiontheory}.
\ngr{This section label doesn't exist. Should it point to methods.detailed?}
There is a large history of literature applying these ideas to obtain measures of the value of forecasts that are tied to a decision making context, primarily in fields such as economics and finance, supply chain management, and meteorology. We review this work only briefly here, and we refer the reader to \cite{yardley2021utility_cost_forecasts} for a general overview, and to \cite{pesaran2002decision_based_eval} and \elr{TODO: identify relevant review or book style summary for meteorology} for discussions focused on applications to economics and meteorology, respectively. In finance, the value of forecasts can often be measured by the profits generated by trading decisions informed by the forecasts, perhaps adjusted for risk levels \cite[e.g.,][]{leitch1991economicForecastEval, cenesizoglu2012returnPredictionEconValue}. In applications to supply chain management and meteorology, the value of forecasts has typically been operationalized by considering the costs associated with decisions regarding the amount of inventory to hold or the level of protection against the impacts of extreme weather events to enact {Peter Catt (2007), Fotios Petropoulos and colleagues (2019), Nada Sanders and Gregory Graman (2009), T.N. Palmer (2002), Florian Pappenberger and colleagues (2015)}. For example, in supply chain management these decisions may incur costs related to holding inventory, labor, or providing poor service, while in meteorology we may need to balance the costs of implementing protective measures with the costs of potentially preventable weather damages. In this framework, a forecast has value if it leads to decisions with low total costs. In all of these fields, analyses have consistently found that common measures of statistical forecast accuracy do not necessarily correspond directly to measures of the value of forecasts as an input to decision making \cite[e.g.,][]{leitch1991economicForecastEval, cenesizoglu2012returnPredictionEconValue}.

>>>>

Here we pursue one possible tether between multivariate forecasts and policy: the allocation of limited medical resources in response to COVID-19 hospitalizations in various regions so as to minimize expected unmet need. Given probabilistic forecasts of hospitalizations in each region, we formulate an allocation algorithm following techniques developed in operations research. We then score forecasts according to how much unmet need their associated allocations would have allowed. We illustrate this scheme with quantile forecasts of COVID-19 hospitalizations in the US at the state level that are recorded in the COVID-19 Forecast Hub, with the goal of determining the allocation of a hypothetical limited resource across the states. The forecast skill ranking given by this allocation scoring rule can vary substantially from the ranking given by the weighted interval score now used by the CDC, especially during surges in hospitalizations such as in late 2021 as the Omicron wave began. We see this as strong evidence that the allocation scoring rule detects forecast value that is missed by traditional accuracy measures and that the general strategy of designing scoring rules that are directly linked to policy performance is a promising research direction for epidemic forecast evaluation.

@@

Infectious disease forecasting models have emerged as important tools in public health. Predictions of disease dynamics have been used to inform decision making about a wide variety of measures to reduce disease spread and/or mitigate the severity of disease outcomes. For example, estimates of expected onset of flu season have been used to aid national vaccination strategies \citep{igboh2023timing}, and forecasts of Ebola dynamics have been used to allocate surveillance resources \citep{meltzer2014estimating, rainisch2015regional}. \cite{bertsimas2021predictionsCOVID} developed tools to inform decision making from infectious disease forecasts, which have been used to inform allocation of limited medical supplies such as ventilators, ICU capacity planning, and vaccine distribution strategy. Models developed by \cite{fox_real-time_2022} have been used to inform resource and care site planning, as well as community guidelines for masking, traveling, dining and shopping (University of Texas, 2022)\nocite{utnews2022}. In April of 2022, the Centers for Disease Control and Prevention (CDC) announced the launch of the Center for Forecasting and Outbreak Analytics (CFA) to translate disease forecasts into decision making (CDC, 2022)\nocite{cdc2022cfa}, indicating that this has been identified as an important direction at the highest levels of government public health response. The value of infectious disease forecasts has typically been measured by how closely they predict disease outcomes such as cases, hospitalizations or deaths using, for example, root mean square error (RMSE) \citep{papastefanopoulos2020covid} or weighted interval score (WIS) \citep{bracher2021evaluating}, however, recently authors have been calling for evaluating forecasts through their impact on policy \citep{marshall2023predictions}.

In decision making settings where it is possible to quantify the utility or loss associated with a particular action, standard tools of decision theory provide a procedure for developing forecast scoring rules that measure the value of forecasts through the quality of the decisions that they lead to. We give an overview of these procedures in Section \ref{sec:methods.detailed.decisiontheory}.
There is a large history of literature applying these ideas to obtain measures of the value of forecasts that are tied to a decision making context, primarily in fields such as economics and finance, supply chain management, and meteorology. We review this work only briefly here, and we refer the reader to \cite{yardley2021utility_cost_forecasts} for a general overview, and to \cite{pesaran2002decision_based_eval} and \cite{murphy1993whatisagoodforecast} for discussions focused on applications to economics and meteorology, respectively. In finance, the value of forecasts can often be measured by the profits generated by trading decisions informed by the forecasts, perhaps adjusted for risk levels \cite[e.g.,][]{leitch1991economicForecastEval, cenesizoglu2012returnPredictionEconValue}. In applications to supply chain management and meteorology, the value of forecasts has typically been operationalized by considering the costs associated with decisions regarding the amount of inventory to hold or the level of protection against the impacts of extreme weather events to enact \cite[e.g.,][]{catt2007assessingcostofforecasterror, petropoulos2019inventoryperformanceforecasting, palmer2002economic, pappenberger2015monetarybenefitfloodwarnings}. For example, in supply chain management these decisions may incur costs related to holding inventory, labor, or providing poor service, while in meteorology we may need to balance the costs of implementing protective measures with the costs of potentially preventable weather damages. In this framework, a forecast has value if it leads to decisions with low total costs. In all of these fields, analyses have consistently found that common measures of statistical forecast accuracy do not necessarily correspond directly to measures of the value of forecasts as an input to decision making \cite[e.g.,][]{leitch1991economicForecastEval, murphy1993whatisagoodforecast, cenesizoglu2012returnPredictionEconValue}.
